%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% use the starred version of the "acknowledgements" environment
%% to omit signatures from this section, e.g.:
%% \begin{acknowledgements*} ... \end{acknowledgements*}
%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin*{acknowledgements}
\chapter*{Acknowledgments}
	Firstly, I  would like to sincerely thank my graduate advisor Frank Rosenbauer for the opportunity to join his laboratory, to contribute to challenging projects and to become acquainted with a broad range of scientific methods. I gratefully acknowledge the possibility to follow my curiosity and address own hypotheses, while being provided guidance when needed.  Furthermore, I much cherish the possibility to attend the SPP1463 meetings in Freiburg (2013) and Menaggio (2016) as well as the hematological symposium in Rotterdam (2015).
	
	I appreciate Joachim Kurtz, Martin Dugas and Carsten Müller-Tidow serving on my thesis committee and providing helpful scientific advice. Sincere thank is given to all collaborators, in particular Michael Rehli, Claudia Gebhard, Frank Lyko, Günther Raddatz and Daniel Lipka. I thankfully acknowledge Ido Amit, Assaf Weiner and Seung-Tae Lee for kindly providing additional data related to their publications upon request.  
	
	The helpful and friendly atmosphere at the institute was highly valuable and I would like to thank all former and current colleagues of the IMTB. There are too many to name them all, but Irina Savelyeva, Verena Gröning, Thorsten König and  Lena Tepe shall be mentioned for strongly supporting the research presented herein.
	
	Pivotal support to this project was also provided by collaborating bioinformaticians and members of online communities. Many thanks to Carolin Walter, Eduard Szöcs, Christian Rohde for teaching me skills and helping with coding questions. I appreciate the whole online community for contributing to free and open-source software and for writing useful tutorials, manuals, forum posts, blog entries and tweets.   
	
	I am fortunate to have been mentored by Elisa Franz while writing this thesis. Many thanks also to Uri Alon, Randy Pausch, Jennifer Heemstra and Ben Barres, whose inspirational talks and encouraging writings on science, motivation and (career) guidance are preeminent. In this context, I also would like to acknowledge Jonathan Haidt and Dan Ariely, whose publications and talks have been extremely helpful to me.  
	
   Last but not least, I am deeply grateful for my family, teachers and friends, who supported me and fostered my passion for science. 

%\end*{acknowledgements}