\chapter{Chromatin}
\minitoc


nice review \cite{Gibcus2013}

chromosome domains \cite{Sexton2015}

\section{TADs}
\label{chap:i:chromatin:tads}

Nanopore based conformations, able to distinguish random from productive enhancer interactions\cite{Allahyar2018}.
\ref{chap:i:enhancers:regulatorygenomics}

Chromosome territories \cite{Bolzer2005} in human diploid fibroblasts. higher-order chromatin arrangements 

TADs\cite{Dixon2012,Hou2012,Nora2012,Sexton2012}


Control of chromatin structure by long noncoding RNA \cite{Boehmdorfer2015}

\cite{Branco2006}

Effect of TADs and TAD formation with regard to enhancers \cite{Long2016}
enhancers in regulation of accessibility \cite{Sleckman1996}

%% key points of \ cite{Ciabrelli2015} 
Metazoan genomes are highly organized inside the cell nucleus.
TADs do not explain alone their spatial organization.
The chromatin shapes TAD's structure and drives its nuclear positioning and function.
Genome-wide association studies revealed four chromatin types.
Main three-dimensional features of each chromatin type, their relationships with TAD organization and epigenetic memory are described.
%%%%


non-coding RNA
He notes one example of an RNA molecule called firre that tethers separate chromosomes together. firre suggests there may be a higher dimension framework in our cells to either turn genes on or off in various contexts. 



the three dimensional cancer genome\cite{Corces2016a}

Review on establishment of chromatin structure\cite{Hug2018}